---
title: "metabinR Vignette"
author: 
-   name: "Anestis Gkanogiannis"
    email: anestis@gkanogiannis.com
package: metabinR
output: 
    BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{metabinR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# About metabinR

description

# Installation

To install `metabinR` package:
```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("metabinR")
```

# Preparation

## Allocate RAM and load required libraries
In order to allocate RAM, a special parameter needs to be passed while JVM 
initializes. JVM parameters can be passed by setting `java.parameters` option.
The `-Xmx` parameter, followed (without space) by an integer value and a 
letter, is used to tell JVM what is the maximum amount of heap RAM that it can
use. The letter in the parameter (uppercase or lowercase), indicates RAM units.
For example, parameters `-Xmx1024m` or `-Xmx1024M` or `-Xmx1g` or `-Xmx1G`, all
allocate 1 Gigabyte or 1024 Megabytes of maximum RAM for JVM.

```{r, eval=TRUE, message=FALSE}
options(java.parameters="-Xmx1G")
unloadNamespace("metabinR")
library(metabinR)
library(utils)
library(BiocFileCache)
```


# Session Info
```{r setup}
utils::sessionInfo()
```
